A thermodynamic view of dusty protoplanetary disks

### 2017-07-01 model.tex
deleted original section two; remixed original section 5 and 6 (now 4 and 5) so the stability discussion goes from general to special, from not needing explicit background to explicit dust distributions

The disk's own gravitational potential $\psi$ is included for
completeness, but we will mostly neglect it unless stated
otherwise. In such cases $\psi$ satisfies the Poisson equation 
\begin{align}
  \nabla^2\psi = 4 \pi G \rho.
\end{align}

@@

\subsection{Entropy of an isothermal dusty gas }

@@

@@

\section{Dynamical equilibria}\label{eqm}

For a given distribution of the dust-fraction $\tepsilon$ (or
dust-to-gas ratio $\epsilon$), the 
mass and momentum Eqs. \ref{masseq}---\ref{momeq} admit     
axisymmetric steady states with $\rho(r,z)$ and 
$\bm{v}=r\Omega(r,z)\hat{\bm{\phi}}$ where $\Omega = v_\phi/r$, which satisfy 
\begin{align}
  r\Omega^2 &= \frac{\p \Phi}{\p r} + \frac{1}{\rho}\frac{\p P}{\p
    r},\label{steady_momr}\\
  0 & = \frac{\p\Phi}{\p z} + \frac{1}{\rho}\frac{\p P}{\p z},\label{steady_momz}
%  0 & = \nabla\cdot\left(\tepsilon\tstop\nabla P\right) \label{steady_dust}
\end{align}
with $P=P(\tepsilon,\rho)$ given by the equation of state
(Eq. \ref{eos}). An explicit solution is presented in
\S\ref{steady_state}.  
However, disk structures obtained this way generally do not satisfy 
the steady state effective energy Eq. \ref{eff_energy} because typically 
$\mathcal{C}\neq0$.  
%for general solutions to 
%Eq. \ref{steady_momr}---\ref{steady_momz}.

% Dust particles settle to the disk midplane on a timescale 
% $t_\mathrm{settle}\sim 1/\OmK^2\tstop$, although for tightly-coupled dust
% particles this is much longer than that to establish vertical
% hydrostatic equilibrium in the gas component ($\sim 1/\OmK$).  
% True steady states may be calculated by invoking an underlying gas
% turbulence, which leads to dust diffusion counter-acting vertical
% settling \citep{takeuchi02, youdin07}. This complication is beyond the
% scope of this paper.

In order to have a well-defined stability problem, we formally
consider equilibria with $\mathcal{C}\equiv 0$. Namely      
\begin{inparaenum}[1)] 
\item 
  unstratified or 2D disks with finite dust-gas drag; %(\S\ref{si});   
\item 
  perfectly-coupled dust with $\tstop=0$. %(\S\ref{results}). 
\end{inparaenum} 
Alternatively, we are assuming that $\tstop$  is sufficiently small
such that the evolution of the background disk (e.g. dust-settling) 
occur on much longer timescales than the 
instabilities of interest.

%Eq. \ref{steady_dust} makes it difficult to obtain equilibrium
%solutions explicitly. However, if the diffusive process is slow and
%can be neglected, then one may just solve
%Eq. \ref{steady_momr}---\ref{steady_momz} with a prescribed (initial)
%distribution of the dust fraction $\tepsilon(r,z)$.

\subsection{Disk structure with a prescribed dust distribution}\label{steady_state}  
For numerical computations, we assume a Gaussian profile in the
dust-to-gas ratio,    
\begin{align}\label{dust_gauss}
  \epsilon(r,z) = \epsilon_0(r)
  \exp{\left[-\frac{z^2}{2\Htilde^2(r)}\right]}. 
\end{align}
%For simplicity we assume the dust-to-gas ratio at the mid-plane
%$\epsilon_0$, as well as its characteristic scale-height
%$\widetilde{H}$, are both constant.
% We take the mid-plane dust-to-gas ratio
%to be a power-law in radius,  
%\begin{align}
%  \epsilon_0(r) = \epsilon_{00}\left(\frac{r}{r_0}\right)^{-d},  
%\end{align}
%where $\epsilon_{00}$ is the dust-to-gas ratio at the fiducial radius
%$r_0$.  
Inserting Eq. \ref{dust_gauss} into vertical hydrostatic equilibrium,
Eq. \ref{steady_momz} and integrating with the approximate
gravitational potential (Eq. \ref{thin_disk_potential}) we obtain the
gas density as
\begin{align}
  &\rhog(r,z)= \notag\\
&\rho_\mathrm{g0}(r)\exp{\left\{ - \frac{z^2}{2\Hgas^2}
    -\epsilon_0\frac{\Htilde^2}{\Hgas^2}\left[1 -
      \exp{\left(-\frac{z^2}{2\Htilde^2}\right)}\right] \right\}}, 
\end{align}
where
\begin{align}
  \Hgas = \frac{c_s}{\OmK}, \quad \OmK \equiv \sqrt{\frac{GM_*}{r^3}},   
\end{align}
is the gas scale-height in the dust-free limit and $\OmK$ is the
Keplerian frequency, respectively.

We usually consider gas-dominated disks with $\epsilon_0 < 1$.  
Then $\rhog(r,z)$ is effectively Gaussian, as in the 
dust-free case. The dust density is approximately 
\begin{align}
  \rhod \simeq \epsilon_0\rho_\mathrm{g0}(r) \exp
        {\left(-\frac{z^2}{2H_\mathrm{d}^2}\right)}, 
\end{align}
with 
\begin{align}
  \frac{1}{H_\mathrm{d}^2} = \frac{1}{\Htilde^2} + \frac{1}{\Hgas^2}, 
\end{align}
and $H_\mathrm{d}$ is the dust-scale height. In numerical
calculations, we  we specify $H_\mathrm{d}< \Hgas$ to obtain 
$\Htilde$ for input.

Finally, we define 
\begin{align}
  Z \equiv \epsilon_0\frac{\Hd}{\Hg} \simeq
  \frac{\Sigma_\mathrm{d}}{\Sigma_\mathrm{g}} 
\end{align}
as a measure of the local metalicity, where $\Sigma_\mathrm{d}$ and
$\Sigma_\mathrm{g}$ are the dust and gas surface densities,
respectively. The second equality holds for $\epsilon_0\ll1$.

%\Sigma_\mathrm{d}$ and the gas
%surface density $\Sigma_\mathrm{g}$.

\subsection{Orbital frequency} 
In the thin-disk approximation, the disk orbital frequency is 
\begin{align}
  \Omega(r,z) = \OmK(r)\left[1 - \frac{3}{2}\frac{z^2}{r^2} +
    \frac{h_\mathrm{g}^2}{\left(1+\epsilon\right)}\frac{\p}{\p\ln{r}}\ln{\left(c_s^2\rhog\right)}
    \right]^{1/2}, 
\end{align}
where 
\begin{align}
  h_\mathrm{g} \equiv \frac{\Hgas}{r}
\end{align}
is the characteristic disk aspect-ratio.

\subsection{Vertical shear}\label{vertshear}
The mixture possess vertical shear. To see this, we eliminate $\Phi$
between Eq. \ref{steady_momr}---\ref{steady_momz} to 
obtain 
\begin{align}\label{vshear}
  r\frac{\p \Omega^2}{\p z} 
%&= \frac{\p\ln{\rho}}{\p r}\frac{\p}{\p
%    z}\left[c_s^2(1-\tepsilon)\right] - \frac{\p\ln{\rho}}{\p z}
%  \frac{\p}{\p r} \left[c_s^2(1-\tepsilon)\right]\\  
   = \frac{1}{\rho}\left(\frac{\p P}{\p r}\frac{\p \seff}{\p z} -\frac{\p
    P}{\p z}\frac{\p \seff}{\p r} \right). 
\end{align}
Writing Eq. \ref{vshear} in terms of the gas density and dust-to-gas
ratio with a power-law temperature profile (Eq. \ref{power_temp}) gives 
\begin{align}\label{vshear2}
  &r\frac{\p \Omega^2}{\p z}  =
  \frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\left\{
  \frac{\p\epsilon}{\p r}\frac{\p\ln{\rhog}}{\p z}
%  -\frac{\p\epsilon}{\p z}\left[\frac{\p\ln{\rhog}}{\p r} + \frac{q}{r}\right]\right.\notag\\
 -\frac{\p\epsilon}{\p z}\frac{\p\ln{P}}{\p r}\right.\notag\\
  &\phantom{ r\frac{\p \Omega^2}{\p z}  =
    \frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\left\{\right\} }
  \left. -\frac{q}{r} \left(1+\epsilon\right)\frac{\p\ln{\rhog}}{\p z}
  \right\}. 
\end{align}
The first two terms correspond to vertical shear caused by spatial
variations in the dust-to-gas ratio. The third term
proportional to $q$ corresponds to vertical shear due to the 
radial temperature gradient, and survives in the dust-free limit.

We can compare these sources by 
evaluating them using the equilibrium
solutions in \S\ref{steady_state}. We assume the disk is radially
smooth so that $\p_r\sim 1/r$. % and the dust-to-gas ratio
% $\epsilon\ll1$. 
This gives

\begin{align}\label{vshear_split}
  \frac{\left|r\p_z\Omega\right|_{\text{
        dust/gas gradient}}}{\left|r\p_z\Omega\right|_{\text{
        temp. gradient}}} \sim
 % \epsilon \frac{\mathrm{max}\left(\delta^2,
  %  1\right)}{\left|q\right|\left(\epsilon + \delta^2\right)},
 \epsilon \frac{\mathrm{max}\left(\delta^2,
    1\right)}{\left|q\right|\left(1+\epsilon\right)\delta^2},
\end{align}
where $\delta\equiv \Htilde/\Hgas$. 
Since $|q|=O(1)$ in PPDs, Eq. \ref{vshear_split} indicates that
vertical shear due to variations in the dust-to-gas ratio dominates 
over that due to the radial temperature gradient for thin dust layers
such that $\delta^2\ll \epsilon$. Otherwise, vertical shear is
associated with the radial temperature gradient. %However, axisymmtric
%\emph{instability} always caused by $\p_rc_s^2$.

\subsection{Dusty vertical buoyancy}\label{vbuoyancy}
Having identified the entropy of an isothermal dusty gas, we find the
vertical buoyancy frequency $N_z$ of the mixture is given by 
\begin{align}
  N_z^2 \equiv - \frac{1}{\rho}\frac{\p P}{\p z}\frac{\p \seff}{\p z} &=
  \frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\frac{\p\ln\rhog}{\p 
  z}\frac{\p\epsilon}{\p z} \\ &
                                  =
  \frac{\epsilon}{\left(1+\epsilon\right)^2}\left(\frac{z}{\Htilde}\right)^2\OmK^2\notag,  
\end{align}
where the second equality assumes $c_s=c_s(r)$ and the final equality
applies to the equilibria in \S\ref{eqm}. Thus,  
\begin{align*}
N_z\lesssim
O\left(\sqrt{\epsilon_0}\OmK\right). 
\end{align*}
However, for well-mixed dust layers such that $H_\epsilon \gg \Hg$, 
$\mathrm{max}\left(N_z\right)$ may occur outside a finite vertical domain.

The physical origin for this vertical buoyancy, which exists even in vertically isothermal
disks, is that coupling gas to dust particles increases
the fluid's inertia, but pressure (i.e. restoring) forces are
unchanged. The increased weight of the fluid resists vertical
oscillations and hence there is an associated buoyancy force. However,
if $\tstop\neq0$ so the gas-dust coupling is imperfect, then gas is no
longer `weighed down' by the dust, and can slip past it. Thus finite
drag diminishes the dust-induced buoyancy. This is similar to reducing
gas buoyancy through cooling \citep{lin15}.

@@

\section{Limiting behaviours}\label{limits}

In this section we discuss the basic properties of the 
dusty-gas mixture using Eqs.\ref{masseq}--\ref{momeq} with 
Eq. \ref{eff_energy} in place of the dust continuity
equation. Self-gravity is neglected here. Our discussion is based on
the corresponding axisymmetric linearized equations. A variable $X$ is
subject to Eulerian perturbations of the form  
\begin{align}
 \real\left[ \delta X(r,z)\exp{\left(-\ii\sigma t\right)}\right], 
\end{align}
where $\sigma$ is the complex mode frequency. For later convenience we
write 
\begin{align}
  \sigma = \ii s - \omega,
\end{align}
where $s$ and $\omega$ are the growth rate and real frequencies,
respectively. Then perturbations have time dependence $e^{st +
  \ii\omega t}$.  Thus for $\omega>0$, perturbations rotate
anti-clockwise in the complex plane.

In Appendix
\ref{var_prin} we derive the integral relation,  
\begin{align}
%&  \sigma^2\int\rho\left(|\dd v_r|^2 + |\dd v_z|^2\right)dV \notag\\
  \sigma^2\mathcal{I}^2
&= \int\left[ \rho
  |\dd v_r|^2A + \rho  \left(\dd v_z \dd v_r^* + \dd v_z^*\dd v_r\right) B +
  \rho |\dd v_z|^2 D\phantom{\frac{1}{1}}\right. \notag\\
&\phantom{===}  \left. + \frac{1}{P}\Bigl\lvert \nabla\cdot\left(P\dd
  \bm{v}\right)\Bigr\rvert^2\right]dV  -\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV \notag\\
&\phantom{===}
- \int P
  \left(\nabla\cdot\dd\bm{v}^*\right)\left(\dd\bm{v}\cdot\nabla\ln{c_s^2}\right)dV,\label{int_rel}
\end{align} 
%The real integral $\mathcal{I}^2>0$ 
where
\begin{align}
  \mathcal{I}^2 \equiv \int\rho\left(|\dd v_r|^2 + |\dd v_z|^2\right)dV
\end{align}
is the meridional kinetic energy, 
and coefficients $A,B,D$ be
read off Eq. \ref{integral_ex}. %Note that $B=C$ from dynamical equilibrium
%(Eq. \ref{vshear}). 
We now consider various limits of Eq. \ref{int_rel}.

@@

In this case, the last two integrals in Eq. \ref{int_rel} vanish. 
Then the condition for axisymmtric \emph{stability}, $\sigma^2>0$, 
is met if the integrand cofficients satisfy $A+D>0$ and $AD-B^2>0$
\citep[][Section 11.6]{ogilvie16}, which becomes 
\begin{align}
  \kappa^2 + c_s^2 \nabla\ln{\rhog}\cdot\nabla\tepsilon &> 0,\label{dusty_solberg1}  \\
  -c_s^2\frac{\p\ln{\rhog}}{\p z}\left(-\kappa^2\frac{\p\tepsilon}{\p
    z} + r\frac{\p\Omega^2}{\p z}\frac{\p \tepsilon}{\p r} \right) & > 0, \label{dusty_solberg2}
\end{align} 
%\begin{align}
%  \kappa^2 - \frac{1}{\rho}\nabla P \cdot \nabla s &> 0, \label{dusty_solberg1}    \\
%  -\frac{1}{\rho}\frac{\p P}{\p z} \left(\kappa^2 \frac{\p s}{\p z} -
%  r\frac{\p\Omega^2}{\p z}\frac{\p s}{\p r}\right) &>0, \label{dusty_solberg2}
%\end{align} 
where %$s = \ln{P/\rho}$, 
$\kappa^2 \equiv r^{-3}\p_r\left(r^4\Omega^2\right)>0$ is the 
square of the epicylic frequency. %We assume $\kappa^2>0$.  
Eq. \ref{dusty_solberg1}---\ref{dusty_solberg2} are in fact the 
the Solberg-Hoiland criteria for axisymmetric stability of adiabatic
gas if one inserts $C_p\ln{\left(1 -
    \tepsilon\right)}$ for the entropy into the standard
expression for said criteria \citep{tassoul78}.

We expect Eq. \ref{dusty_solberg1} to be satisfied in typical
protoplanetary disks where the dust-to-gas ratio increases in the same
direction as the local pressure gradient, which is equivalent to the
gas density gradient for isothermal gas. Thus for 
instability we need to violate Eq. \ref{dusty_solberg2}. We explore
this below.

\subsubsection{Vertically well-mixed dust}\label{vert_mixed_theory}
In the special case where the dust-to-gas ratio is
uniform in $z$ the left-hand-side of Eq. \ref{dusty_solberg2} becomes
\begin{align}
-\frac{c_s^4}{(1+\tepsilon)^4}\left(\frac{\p\ln{\rhog}}{\p
    z}\right)^2\left(\frac{\p\epsilon}{\p r}\right)^2<0.
\end{align}
Thus the disk is \emph{unstable}, because there is no effective vertical 
buoyancy to stabilize vertical motions ($N_z=0$). This situation is 
identical to the pure gas, adiabtic simulations of \cite{nelson13}
where the gas entropy is vertically uniform but has a radial
gradient. The authors indeed find instability. We give a numerical
example in the dusty context in \S\ref{vert_mixed}.

\subsubsection{Application to protoplanetary disks} 
%In order to apply the stability criteria
%Eq. \ref{dusty_solberg1}---\ref{dusty_solberg2} to typical
%protoplanetary disks 
We now consider the specific equilibria described in 
\S\ref{eqm}. We assume the disk is approximately Keplerian so that
$\kappa\simeq\OmK$, and that the vertical gas distribution is   
Gaussian. In terms of the dust-to-gas
ratio $\epsilon$, Eq. \ref{dusty_solberg2}
becomes 
\begin{align}
  % &1 + \frac{\epsilon}{\left(1+
  %   \epsilon\right)^2}\left(h_\mathrm{g}^2\frac{\p\ln{\rhog}}{\p\ln{r}}
  % \frac{\p\ln{\epsilon}}{\p\ln{r}} + 
  % \frac{z^2}{\Htilde^2}\right)>0 \label{stability_est1},   \\ 
&1 - \frac{\epsilon
  h_\mathrm{g}^2}{\left(1+\epsilon\right)^2}
  \frac{\p\ln{\epsilon}}{\p\ln{r}}\left(-\frac{\p\ln{\rhog}}{\p\ln{r}}+\frac{\Htilde^2}{H_\mathrm{g}^2}\frac{\p\ln{\epsilon}}{\p\ln{r}}\right)
  > 0 \label{stability_est2},
\end{align}
for \emph{stability}. The second term is usually stabilizing in 
PPDs for reasons given in \S\ref{iso_perfect}. The third term is
always destabilizing, but is small in smooth, thin disks with radial
gradients $O(1/r)$, $h_\mathrm{g}\ll 1$, 
provided that $H_\epsilon/\Hg$ is not large (e.g. some dust settling
has occured). This means that typical dusty PPDs are stable to
axisymmetric perturbations, a conclusion valid for arbitrarily thin
dust layers.

%%%%%%%%%%%%%%%%%%

To violate Eq. \ref{stability_est2} and obtain instability, notice the 
LHS is a quadratic in $\p_r\epsilon$. Thus instability is possible for
sufficiently large (in magnitude) radial gradients  in the dust-to-gas
ratio,  
\begin{align}
  \frac{\p\ln{\epsilon}}{\p\ln{r}} > \chi_+ \quad \mathrm{or} \quad 
  \frac{\p\ln{\epsilon}}{\p\ln{r}} < \chi_-,
\end{align}
for \emph{instability}, where
\begin{align}\label{spm}
\chi_\pm = \frac{1}{2}\frac{H_\mathrm{g}^2}{\Htilde^2} 
  \left[
  \frac{\p\ln{\rhog}}{\p\ln{r}} \pm 
  \sqrt{
  \left(\frac{\p\ln{\rhog}}{\p\ln{r}}\right)^2 + 
  4 \frac{\Htilde^2}{H_\mathrm{g}^2}
  \frac{\left(1+\epsilon\right)^2}{\epsilon h_\mathrm{g}^2}
  }
  \,\right]. 
\end{align} 
In typical accretion disks where $\p_r\rhog<0$, instability is easier
for increasing dust-to-gas ratios ($\p_r\epsilon > 0$).

We can neglect density gradients in Eq. \ref{spm} 
if $r\p_r\ln{\rhog}\sim O(1)$ and $H_\epsilon/\Hgas\gg
\sqrt{\epsilon}\hgas$. For example, if $\epsilon\simeq 0.01$ and
$\hgas\simeq 0.05$, then we require $H_\epsilon/\Hg\gg
5\times10^{-3}$. (So the dust layer thickness $\Hd\gg 5
\times10^{-3}\Hgas$ as well.) Then 
instability requires 
\begin{align}\label{dust_edge}
\left|\frac{\p\ln{\epsilon}}{\p r}\right| \gtrsim
  \frac{1}{\Htilde}\frac{\left(1+\epsilon\right)}{\sqrt{\epsilon}}.
%\simeq
 % \frac{1}{\sqrt{\epsilon}H_\epsilon}, 
\end{align}
%where the second equality applies to $\epsilon\ll 1$. 
That is, if the radial lengthscale of the dust-to-gas ratio,
$L_\epsilon\ll O(\Htilde)$, i.e. its vertical lengthscale,  
then the system is potentially unstable.  
Taking $\epsilon\sim 0.01$,  we find that 
for thin dust layers with $H_\epsilon \simeq \Hd\ll \Hg$,  instability
requires $L_\epsilon\ll O(0.1\Hg)$, i.e. the dust-to-gas ratio must 
vary on an extremely short lengthscale. This might be achieved, for
example, at sharp edges associated with gaps opened by giant planets. 
{\bf mention this is equivalent to sharp entropy gradients?}

>>>>

The disk's own gravitational potential $\psi$ is included for
completeness, but we will mostly neglect it unless stated
otherwise. In such cases $\psi$ satisfies the Poisson equation 
\begin{align}
  \nabla^2\psi = 4 \pi G \rho.\label{poisson}
\end{align}

@@

\subsection{Entropy and buoyancy of isothermal dusty gas }\label{dusty_entropy}

@@

%\subsection{Dusty buoyancy forces}
Having identified the entropy of an isothermal dusty gas, we can also
define the disk vertical buoyancy frequency $N_z$ as   
%$N_z$ of the mixture is given by 
\begin{align}
  N_z^2 \equiv - \frac{1}{\rho}\frac{\p P}{\p z}\frac{\p \seff}{\p z}
  =c_s^2(r) \frac{\p \ln{\rhog}}{\p z}\frac{\p \tepsilon}{\p z},
%  \quad 
 % N_z^2 \equiv - \frac{1}{\rho}\frac{\p P}{\p z}\frac{\p \seff}{\p z}.
%&=
  %\frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\frac{\p\ln\rhog}{\p 
%  z}\frac{\p\epsilon}{\p z} \\ &
  %                                =
 % \frac{\epsilon}{\left(1+\epsilon\right)^2}\left(\frac{z}{\Htilde}\right)^2\OmK^2\notag,  
\end{align}
%Typically in a thin, smooth disk we have $|N_r|\ll |N_z|$. 
where the second equality assumes a vertically isothermal disk. A
similar definition exists for the radial buoyancy frequency
$N_r$. Typically in a thin, smooth disk we have $|N_r|\ll |N_z|$.

A vertical buoyancy force exists even in vertically isothermal
dusty disks, because coupling gas to dust particles increases
the fluid's inertia, but pressure (i.e. restoring) forces are
unchanged. The increased weight of the fluid resists vertical
oscillations and hence there is an associated buoyancy force. However,
if $\tstop\neq0$ so the gas-dust coupling is imperfect, then gas is no
longer `weighed down' by the dust, and can slip past it. Thus finite
drag diminishes the dust-induced buoyancy. This is similar to reducing
gas buoyancy through cooling \citep{lin15}.

@@

\section{General stability criteria for dusty gas}\label{limits}

In this section we discuss the stability properties of the 
dusty-gas mixture using a variational principle. This method does not
require explicit solutions to the equilibrium equations. Self-gravity is neglected here.

\subsection{Steady states}\label{eqm}
We consider axisymmetric steady states as background equilibria to
perturb. For a given distribution of the dust-fraction $\tepsilon$ (or
dust-to-gas ratio $\epsilon$), the 
mass and momentum Eqs. \ref{masseq}---\ref{momeq} admit     
axisymmetric steady states with $\rho(r,z)$ and 
$\bm{v}=r\Omega(r,z)\hat{\bm{\phi}}$ where $\Omega = v_\phi/r$, which satisfy 
\begin{align}
  r\Omega^2 &= \frac{\p \Phi}{\p r} + \frac{1}{\rho}\frac{\p P}{\p
    r},\label{steady_momr}\\
  0 & = \frac{\p\Phi}{\p z} + \frac{1}{\rho}\frac{\p P}{\p z},\label{steady_momz}
%  0 & = \nabla\cdot\left(\tepsilon\tstop\nabla P\right) \label{steady_dust}
\end{align}
with $P=P(\tepsilon,\rho)$ given by the equation of state
(Eq. \ref{eos}). An explicit solution is presented in 
\S\ref{steady_state}, when we explicitly analyse the stability of
protoplanetary disks.

However, disk structures obtained this way generally do not satisfy 
the steady state effective energy Eq. \ref{eff_energy} because typically 
$\mathcal{C}\neq0$.  In order to have a well-defined stability problem, we formally
consider equilibria with $\mathcal{C}\equiv 0$. Namely      
\begin{inparaenum}[1)] 
\item 
  unstratified or 2D disks with finite dust-gas drag; %(\S\ref{si});   
\item 
  perfectly-coupled dust with $\tstop=0$. %(\S\ref{results}). 
\end{inparaenum} 
Alternatively, we are assuming that $\tstop$  is sufficiently small
such that the evolution of the background disk (e.g. dust-settling) 
occur on much longer timescales than the instabilities of interest.

@@

%\section{Limiting behaviours}\label{limits}

\subsection{Integral relation}

%Our discussion is based on
%the corresponding axisymmetric linearized equations. 
We consider axisymmetric Eulerian perturbations to a variable $X$ of
the form  
\begin{align}
 \real\left[ \delta X(r,z)\exp{\left(-\ii\sigma t\right)}\right], 
\end{align}
where $\sigma$ is the complex mode frequency. For later convenience we
write 
\begin{align}
  \sigma = \ii s - \omega,
\end{align}
where $s$ and $\omega$ are the growth rate and real frequencies,
respectively. Then perturbations have time dependence $e^{st +
  \ii\omega t}$.  Thus for $\omega>0$, perturbations rotate
anti-clockwise in the complex plane.

In Appendix \ref{var_prin} we derive the integral relation from the
corresponding linearized equations (Eqs.\ref{masseq}--\ref{momeq}, and 
Eq. \ref{eff_energy} in place of the dust continuity equation): 
\begin{align}
%&  \sigma^2\int\rho\left(|\dd v_r|^2 + |\dd v_z|^2\right)dV \notag\\
  \sigma^2\mathcal{I}^2
&= \int\left[ \rho
  |\dd v_r|^2A + \rho  \left(\dd v_z \dd v_r^* + \dd v_z^*\dd v_r\right) B +
  \rho |\dd v_z|^2 D\phantom{\frac{1}{1}}\right. \notag\\
&\phantom{===}  \left. + \frac{1}{P}\Bigl\lvert \nabla\cdot\left(P\dd
  \bm{v}\right)\Bigr\rvert^2\right]dV  -\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV \notag\\
&\phantom{===}
- \int P
  \left(\nabla\cdot\dd\bm{v}^*\right)\left(\dd\bm{v}\cdot\nabla\ln{c_s^2}\right)dV,\label{int_rel}
\end{align} 
%The real integral $\mathcal{I}^2>0$ 
where
\begin{align}
  \mathcal{I}^2 \equiv \int\rho\left(|\dd v_r|^2 + |\dd v_z|^2\right)dV
\end{align}
is the meridional kinetic energy, 
and coefficients $A,B,D$ be
read off Eq. \ref{integral_ex}. %Note that $B=C$ from dynamical equilibrium
%(Eq. \ref{vshear}). 
We now consider various limits of Eq. \ref{int_rel}.

@@

In this case, the last two integrals in Eq. \ref{int_rel} vanish. 
Then the condition for axisymmtric \emph{stability}, $\sigma^2>0$, 
is met if the integrand cofficients satisfy $A+D>0$ and $AD-B^2>0$
\citep[][Section 11.6]{ogilvie16}, which becomes 
\begin{align}
  \kappa^2 + c_s^2 \nabla\ln{\rhog}\cdot\nabla\tepsilon &> 0,\label{dusty_solberg1}  \\
  -c_s^2\frac{\p\ln{\rhog}}{\p z}\left(-\kappa^2\frac{\p\tepsilon}{\p
    z} + r\frac{\p\Omega^2}{\p z}\frac{\p \tepsilon}{\p r} \right) & > 0, \label{dusty_solberg2}
\end{align} 
%\begin{align}
%  \kappa^2 - \frac{1}{\rho}\nabla P \cdot \nabla s &> 0, \label{dusty_solberg1}    \\
%  -\frac{1}{\rho}\frac{\p P}{\p z} \left(\kappa^2 \frac{\p s}{\p z} -
%  r\frac{\p\Omega^2}{\p z}\frac{\p s}{\p r}\right) &>0, \label{dusty_solberg2}
%\end{align} 
where %$s = \ln{P/\rho}$, 
$\kappa^2 \equiv r^{-3}\p_r\left(r^4\Omega^2\right)>0$ is the 
square of the epicylic frequency. %We assume $\kappa^2>0$.  
Eq. \ref{dusty_solberg1}---\ref{dusty_solberg2} are in fact the 
the Solberg-Hoiland criteria for axisymmetric stability of adiabatic
gas if one inserts $C_p\ln{\left(1 -
    \tepsilon\right)}$ for the entropy into the standard
expression for said criteria \citep{tassoul78}. 
We expect Eq. \ref{dusty_solberg1} to be satisfied in typical
protoplanetary disks where the dust-to-gas ratio increases in the same
direction as the local pressure gradient, which is equivalent to the
gas density gradient for isothermal gas.

On the other hand, Eq. \ref{dusty_solberg2} can be violated in
disks if the dust is vertically well-mixed ($\p_z\epsilon =
0$) but  radially-varying ($\p_r\epsilon\neq0$). In this case the
left-hand-side of Eq. \ref{dusty_solberg2} becomes
\begin{align}
-\frac{c_s^4}{(1+\tepsilon)^4}\left(\frac{\p\ln{\rhog}}{\p
    z}\right)^2\left(\frac{\p\epsilon}{\p r}\right)^2<0.
\end{align}
Such an isothermal dusty disk is \emph{unstable} because there is no effective vertical 
buoyancy to stabilize vertical motions ($N_z=0$). This situation is 
identical to the pure gas, adiabtic simulations of \cite{nelson13}
where the gas entropy is vertically uniform but has a radial
gradient. The authors indeed find instability. We give a numerical
example in the dusty context in \S\ref{vert_mixed}.
